Product graphics

### 2009-11-26 product-graphics.tex
More work, including start on unit tests.
Starting to thrash the paper into some sort of shape

The Grammar of Graphics [reference] is a grammar describing the components of statistical graphics and how they can be combined together.  These components are very flexible and can be used to describe any type of plot.  However, the book (Grammar with a capital G) does not present many examples of graphics for categorical data, and in those examples the use of the grammar gives little insight into plot.  This paper describes common categorical plots, and presents a specialisation of the Grammar of Graphics specifically designed to describe these types of plots and give insight into the types of relationships that they can show.

@@

Area plots are the graphical equivalent to contingency tables.  And like contingency tables, they can be used to display raw counts, or to display proportions or percentages.  Similarly, it may be useful to augment area plots with marginal displays as we augment contingency tables with marginal sums.

Area plots different to other types of graphics, because must be able to flexible change level of aggregation.  Values are typically counts, but can also be sums of observation level weights.

These techniques are also closely related to TreeMaps and dimensional stacking \citep{leblanc:1990}

What are the common components of all these plots, and how can be describe them in a succinct way that also allows us to describe what tasks each display is good for?  We propose that there is one key feature that makes these plots special.  That is there their coordinate system, which is hierarchical.

In most coordinate systems that we are used to, such as Cartesian or polar coordinates, are symmetric in the sense that you can locate a point by starting from either dimension.  For example, in Cartesian coordinates you can identify either the x- or y-coordinate first, and the other next.  This also implies a distance symmetry when projected onto 2D Euclidean geometry. When you vary one coordinate by a small amount (holding all others constant) points in the projective geometry will be close.

Neither of these properties hold for hierarchical coordinate systems.

We will discuss hierachical coordinate systems in general, and then present a specific hierachical coordinate system which gives rise to the area plots described previously.  We also present an R implementation, in the R package {\tt ggplot}.

Related work:

\citep{vliegen:2006}
\citep{ahlberg:1996}

\citep{hartigan:1984,hartigan:1981}
\citep{friendly:1999,friendly:1994}
\citep{theus:1999}
\citep{hofmann:2000,hofmann:2001,hofmann:2003}

Goals:

@@

\section{Data}
\label{sec:data}

\subsection{Factorising probabilities}

@@

Compare to area of mosaic plot.

One other special 2d case is the equal bin size plot - which all 1d x 1d primitive plots collapse down to when counts are equal.  The important difference is the formatting - a grid rather than nested boxes.

We have assumed the probabilities represent counts, but WLOG we can use any set of non-negative additive weights instead.  Some of the applications of such weighted plots are described in \citet{unwin:2007}.

Binning.

Convention when using binned continuous data is to remove the gap between regions.

@@

Impossible to simultaneously satisfy all constraints.  Every product plot requires the creator to trade off between these.  We will focus our attention on rectangular partitions.  Special cases of $a(i, j) = w(i, j)^x * h(i, j)^y$,   $x + y = 1$.

Restrictions on partitioning

\begin{itemize}
  \item Area should be proportional to weight
  \item Containment
  \item Rectangular
  \item Non-overlapping
\end{itemize}

Pseudo 2d, where you take a long 1d structure and wrap it into 2d.

Pseduo 1d, where you make a new variable from multiple variables.

@@

Basic principle of all area plots is area should be proportional to probability. For the plots we are interested in, the shape used to represent the count is a rectangle, so that $a(i, j) = w(i, j) * h(i, j)$. Each plot type constrains this relationship in a different way:

@@

  \item {\bf treemap}: no restrictions on height and width.

@@

\begin{itemize}
  \item {\bf fluct}: width and height proportional to square root of value. Flucts allow comparisons of (approximately) absolute numbers in two directions.

@@

@@

@@

\subsubsection{Small values and zeros}
\subsubsection{Cascading}

@@

@@

\begin{itemize}
  \item Pie charts
  \item Fourfold displays \citep{friendly:1995}
  \item Bullseye chart
\end{itemize}

@@

@@

@@

Labelling these plots is particularly challenging.

\begin{itemize}
  \item Additional info 
  \begin{itemize}
    \item colour (map of the market)
    \item texture (sieve plots)
    \item embedded plots (time series in lab escape)
    \item photographs
    \item text (tables)
  \end{itemize}

@@

>>>>

% JASA methods
%

% The Grammar of Graphics [reference] describes the independent components of
% statistical graphics and how they can be combined together. These components
% are very flexible and can be used to describe any type of plot. However, the
% book (Grammar with a capital G) does not present many examples of graphics
% for categorical data, and in those examples the use of the grammar gives
% little insight into plot. This paper describes common categorical plots, and
% presents a specialisation of the Grammar of Graphics specifically designed
% to describe these types of plots and give insight into the types of
% relationships that they can show.

@@

@@

Area plots are the graphical equivalent of contingency tables. Like contingency tables, they can be used to display raw counts, or conditioned and marginal breakdowns. Similarly, it may be useful to augment area plots with marginal displays as we augment contingency tables with marginal sums.  We can decompose these graphics in a similar way to the way we decompose products of probabilities.

@@

These techniques are also closely related to dimensional stacking \citep{leblanc:1990}

Generalised treemaps: \citep{vliegen:2006}
Cocktailmaps: \citep{ahlberg:1996}
Mosaic plots: \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003}
Extension to mosaic: \citep{friendly:1999,hofmann:2000}
Connection to models: \citep{theus:1999,hofmann:2001}

% Hierarchical coordinate systems 
%
% What are the common components of all these plots, and how can be describe
% them in a succinct way that also allows us to describe what tasks each
% display is good for?  We propose that there is one key feature that makes
% these plots special.  That is there their coordinate system, which is
% hierarchical.  
%
% In most coordinate systems that we are used to, such as Cartesian or polar
% coordinates, are symmetric in the sense that you can locate a point by
% starting from either dimension.  For example, in Cartesian coordinates you
% can identify either the x- or y-coordinate first, and the other next.  This
% also implies a distance symmetry when projected onto 2D Euclidean geometry.
% When you vary one coordinate by a small amount (holding all others constant)
% points in the projective geometry will be close.
%
% Neither of these properties hold for hierarchical coordinate systems. 
%
% We will discuss hierachical coordinate systems in general, and then present
% a specific hierachical coordinate system which gives rise to the area plots
% described previously.  We also present an R implementation, in the R package
% {\tt ggplot}.

@@

Impossible to simultaneously satisfy all constraints.  Every product plot requires the creator to trade off between these.

Constraints on partitioning

\begin{itemize}
  \item Area should be proportional to weight
  \item Containment / Non-overlapping
\end{itemize}

We will add one more constraint to reduce the space of possible

\begin{itemize}
  \item Rectangular partitions
\end{itemize}

We will focus our attention on rectangular partitions.  In general,  $h(i, j) = c(i, j)^x$ $w(i, j) = c(i, j)^y$, $x + y = 1$.   Basic principle of all area plots is area should be proportional to probability. For the plots we are interested in, the shape used to represent the count is a rectangle, so that $a(i, j) = w(i, j) * h(i, j)$.

@@

Each plot type constrains this relationship in a different way:

@@

  \item {\bf treemap}: no restrictions on height and width.   At first glance, it may seem strange to include the treemap as a 1d primitive, but it lays out a single variable in 2d dimensions, just like the other approaches.

@@

\begin{itemize}
  \item {\bf fluct}: width and height proportional to square root of value. Flucts allow comparisons of absolute numbers in two directions.

@@

One special 2d case is the equal bin size plot - which all 1d x 1d primitives and the fluct collapse to when counts are equal.

@@

Plot templates combine the simple primitives into well-known graphics:

@@

\subsubsection{Area not proportional to weight}

It can be useful to violate the constraint that area should be proportional to weight to distinguish between zeros and very small values.  A zero weight should have zero area, but giving it positive area can be useful so it's actually possible to see!  Similarly for very small values which would otherwise occupy less than a pixel, useful to constain to minimal perceptible size.

Antony's zooming, where you also truncate the maximum size.

Image.

\subsubsection{Overlapping rectangles}

@@

Example

@@

These displays generalise the fourfold displays of \citet{friendly:1995}.

\begin{itemize}
  \item Pie charts
  \item Nightingale
  \item Bullseye chart 
\end{itemize}

@@

\section{Margining}
\label{sec:margining}

Mathematically defining margining operation.

@@

\section{Variations}
\label{sec:variations}

% Pseudo 2d, where you take a long 1d structure and wrap it into 2d.
% 
% Pseduo 1d, where you make a new variable from multiple variables.

We have assumed the probabilities represent counts, but without loss of generality we can use any set of non-negative, additive weights instead.  Some of the applications of such weighted plots are described in \citet{unwin:2007}.  Include a brief summary here.

We distinguish nested data from crossed data by the number of interactions with missing values.  Completely crossed data contains every combination of all variables.  Nested data contains

Aka sparse vs. dense.

It is not always possible to determine nested data from inspection of the data alone, but may need knowledge of the data collection.  For example, if a survey of teachers within schools was performed, and each teacher was identified by a identifier unique within a school, it might look like the data was only missing a few combinations, when in fact it is meaningless.

Or when you have a hierarchy.

Example dataset

Possibilities: create new variable which is interaction of nested vars, or drop missings/zeros.  Treemaps deal almost exclusively with nested data.

@@

Labelling these plots is particularly challenging.  Future work!

\begin{itemize}
  \item Additional info 
  \begin{itemize}
    \item colour (map of the market)
    \item texture (sieve plots)
    \item photographs
    \item text (tables)
    \item embedded plots (time series in lab escape)
  \end{itemize}

@@

\section{Conclusions} % (fold)
\label{sec:conclusions}

% section conclusions (end)
